---
title: "01-eda"
author: "YANKAI MA"
date: "2026-07-19"
output: 
  pdf_document: 
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(corrr)

theme_set(theme_minimal())
```

```{r}
df <- read_csv(
  "data/cleaned_panel_2021_2025.csv",
  col_types = cols(city_code = col_character(), .default = col_guess())
)

# Rename building stat columns
ncol_bldg <- ncol(df) - 6
colnames(df)[4:ncol_bldg] <- c(
  "all_type_building_count_total",
  "all_type_building_area",
  "living_count",
  "living_area",
  "dormitory_count",
  "dormitory_area",
  "mixed_use_count",
  "mixed_use_area",
  "agri_count",
  "agri_area",
  "mining_count",
  "mining_area",
  "plant_count",
  "plant_area",
  "energy_count",
  "energy_area",
  "it_count",
  "it_area",
  "logis_count",
  "logis_area",
  "retail_count",
  "retail_area",
  "finance_count",
  "finance_area",
  "real_estate_count",
  "real_estate_area",
  "hotel_count",
  "hotel_area",
  "education_count",
  "education_area",
  "health_count",
  "health_area",
  "other_service_count",
  "other_service_area",
  "official_count",
  "official_area",
  "other_type_count",
  "other_type_area"
)
glimpse(df)
```

## Predictor: Population Growth Rate

The target predictor variable for supervised learning is `pop_growth_rate` that is included population stat.

### Distributions and Long Tail

Overall, there is negative growth (median of approximately −1.2%, due to a low birth rate and aging population). However, there are long tails on both sides: the maximum value is +22.3%, which, when used as Y, results in high variance.

```{r}
summary(df %>% select(starts_with("pop_")))

p_full <- ggplot(df, aes(pop_growth_rate)) +
  geom_histogram(bins = 80,
                 fill = "steelblue",
                 color = "white") +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "red") +
  labs(x = "Population Growth Rate (%)", y = "Observations (cities * years)", title = "All: Long Tail (20%)")

p_zoom <- ggplot(df, aes(pop_growth_rate)) +
  geom_histogram(bins = 60,
                 fill = "steelblue",
                 color = "white") +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "red") +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(x = "Population Growth Rate (%)", y = "Observations (cities * years)", title = "Main: 5%")

p_full / p_zoom
```

### Verification: The Relationship Between the Variance of the Growth Rate and Population Size

As a matter of common sense, the smaller the population, the more volatile the growth rate tends to be, meaning that even a small number of people moving in or out can cause significant fluctuations. Validate the results using box plots based on population size.

```{r}
df %>%
  filter(pop_population > 0) %>%
  mutate(pop_bin = cut(
    pop_population,
    breaks = c(0, 500, 1000, 5000, 10000, 50000, Inf),
    labels = c("<500", "0.5k-1k", "1k-5k", "5k-10k", "10k-50k", ">50k")
  )) %>%
  ggplot(aes(pop_bin, abs(pop_growth_rate))) +
  geom_boxplot(
    fill = "steelblue",
    outlier.alpha = 0.1,
    outlier.size = 0.5
  ) +
  scale_y_log10() +
  labs(x = "Population Size", y = "|Growth Rate| (%), Log Scale", title = "The variance in change rates for small towns is greater than large")

```

## Data Validation

### Missing values

The number of municipalities should be consistent each year, and no missing values in the data columns.

```{r}
# Balance
df %>% count(year)

# Missing values
df %>%
  select(all_type_building_count_total:other_type_area) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "col", values_to = "n_na") %>%
  filter(n_na > 0)
```

### All-zero rows

```{r}
zero_rows <- df %>% filter(all_type_building_area == 0)

# Each year
zero_rows %>% count(year)
zero_rows %>% distinct(city_code, city_name)
```

## Constructing Scaled Variable

By dividing the area of each subcomponent by the total area, we obtain 18 proportional variables, the sum of which in each row exactly equals 1 (composite data / simplex constraint), for use in modeling.

```{r}
df_prop <- df %>%
  filter(all_type_building_area > 0) %>%
  mutate(
    across(
      ends_with("_area") & !all_type_building_area,
      ~ .x / all_type_building_area,
      .names = "p_{.col}"
    )
  ) %>%
  rename_with( ~ str_remove(., "_area$"), starts_with("p_"))

# Verify simplex
df_prop %>%
  transmute(row_sum = rowSums(across(starts_with("p_")))) %>%
  summarise(min = min(row_sum), max = max(row_sum))
```

## Exploration of correlation

### Screening Using Pearson's Correlation (composite data)

```{r}
# Correlation between proportion of area by industry and change rates
corr_y <- df_prop %>%
  select(starts_with("p_"), pop_growth_rate) %>%
  correlate() %>% # Default use pearson
  focus(pop_growth_rate) %>%
  mutate(term = str_remove(term, "^p_")) %>%
  arrange(pop_growth_rate)

corr_y %>%
  ggplot(aes(fct_reorder(term, pop_growth_rate), pop_growth_rate)) +
  geom_col(aes(fill = pop_growth_rate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("steelblue", "firebrick"),
    labels = c(
      "Negative correlation (population decline)",
      "Positive correlation (population growth)"
    ),
    name = NULL
  ) +
  labs(x = NULL, y = "Pearson's Correlation", title = "Correlation between proportion of area by industry and change rates") +
  theme(legend.position = "bottom")
```
