---
title: "01-eda"
author: "YANKAI MA"
date: "2026-07-19"
output: 
  pdf_document: 
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)

theme_set(theme_minimal())
```

```{r}
df <- read_csv(
  "data/cleaned_panel_2021_2025.csv",
  col_types = cols(city_code = col_character(), .default = col_guess())
)

# Rename building stat columns
ncol_bldg <- ncol(df) - 6
colnames(df)[4:ncol_bldg] <- c(
  "all_type_building_count_total",
  "all_type_building_area",
  "living_count",
  "living_area",
  "dormitory_count",
  "dormitory_area",
  "mixed_use_count",
  "mixed_use_area",
  "agri_count",
  "agri_area",
  "mining_count",
  "mining_area",
  "plant_count",
  "plant_area",
  "energy_count",
  "energy_area",
  "it_count",
  "it_area",
  "logis_count",
  "logis_area",
  "retail_count",
  "retail_area",
  "finance_count",
  "finance_area",
  "real_estate_count",
  "real_estate_area",
  "hotel_count",
  "hotel_area",
  "education_count",
  "education_area",
  "health_count",
  "health_area",
  "other_service_count",
  "other_service_area",
  "official_count",
  "official_area",
  "other_type_count",
  "other_type_area"
)
glimpse(df)
```

## Predictor: Population Growth Rate

The target predictor variable for supervised learning is `pop_growth_rate` that is included population stat.

### Distributions and Long Tail

Overall, there is negative growth (median of approximately −1.2%, due to a low birth rate and aging population). However, there are long tails on both sides: the maximum value is +22.3%, which, when used as Y, results in high variance.

```{r}
summary(df %>% select(starts_with("pop_")))

p_full <- ggplot(df, aes(pop_growth_rate)) +
  geom_histogram(bins = 80,
                 fill = "steelblue",
                 color = "white") +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "red") +
  labs(x = "Population Growth Rate (%)", y = "Observations (cities * years)", title = "All: Long Tail (20%)")

p_zoom <- ggplot(df, aes(pop_growth_rate)) +
  geom_histogram(bins = 60,
                 fill = "steelblue",
                 color = "white") +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "red") +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(x = "Population Growth Rate (%)", y = "Observations (cities * years)", title = "Main: 5%")

p_full / p_zoom
```

### Verification: The Relationship Between the Variance of the Growth Rate and Population Size

As a matter of common sense, the smaller the population, the more volatile the growth rate tends to be, meaning that even a small number of people moving in or out can cause significant fluctuations. Validate the results using box plots based on population size.

```{r}
df %>%
  filter(pop_population > 0) %>%
  mutate(pop_bin = cut(
    pop_population,
    breaks = c(0, 500, 1000, 5000, 10000, 50000, Inf),
    labels = c("<500", "0.5k-1k", "1k-5k", "5k-10k", "10k-50k", ">50k")
  )) %>%
  ggplot(aes(pop_bin, abs(pop_growth_rate))) +
  geom_boxplot(
    fill = "steelblue",
    outlier.alpha = 0.1,
    outlier.size = 0.5
  ) +
  scale_y_log10() +
  labs(x = "Population Size", y = "|Growth Rate| (%), Log Scale", title = "The variance in change rates for small towns is greater than large")

```
