---
title: "02-modeling"
author: "YANKAI MA"
date: "2026-07-22"
output: 
  pdf_document: 
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = "figures/",
  dev = "png",
  dpi = 300
)
library(tidyverse)
library(glmnet)   # lasso
library(broom)    # tidy lm

theme_set(theme_minimal())
```

## Data Preparation

```{r}


df <- read_csv(
  "data/cleaned_panel_2021_2025.csv",
  col_types = cols(city_code = col_character(), .default = col_guess()),
  show_col_types = FALSE
)
ncol_bldg <- ncol(df) - 6
colnames(df)[4:ncol_bldg] <- c(
  "all_type_building_count_total",
  "all_type_building_area",
  "living_count",
  "living_area",
  "dormitory_count",
  "dormitory_area",
  "mixed_use_count",
  "mixed_use_area",
  "agri_count",
  "agri_area",
  "mining_count",
  "mining_area",
  "plant_count",
  "plant_area",
  "energy_count",
  "energy_area",
  "it_count",
  "it_area",
  "logis_count",
  "logis_area",
  "retail_count",
  "retail_area",
  "finance_count",
  "finance_area",
  "real_estate_count",
  "real_estate_area",
  "hotel_count",
  "hotel_area",
  "education_count",
  "education_area",
  "health_count",
  "health_area",
  "other_service_count",
  "other_service_area",
  "official_count",
  "official_area",
  "other_type_count",
  "other_type_area"
)


# Percentage of total area accounted for by the 18 subcategories p
df_prop <- df %>%
  filter(all_type_building_area > 0) %>%            # Drop all zero rows
  mutate(
    across(
      ends_with("_area") & !all_type_building_area,
      ~ .x / all_type_building_area,
      .names = "p_{.col}"
    )
  ) %>%
  rename_with(~ str_remove(., "_area$"), starts_with("p_"))

# Proportions of the 17 subcategories (excluding p_living)
fine_cols <- setdiff(grep("^p_", names(df_prop), value = TRUE), "p_living")
# Full samples
df_full <- df_prop
# Threshold: pop > 10000
df_prop <- df_prop %>% filter(pop_population > 10000)

# 70/30 Split train and test set
set.seed(1)
idx <- sample(nrow(df_prop), 0.7 * nrow(df_prop))
df_train <- df_prop[idx, ]
df_test  <- df_prop[-idx, ]

```

## Explanation for `pop > 10000`

The EDA phase revealed that small villages and towns were outliers and were therefore excluded. The table below directly illustrates the benefits of this approach—after filtering by population threshold, the R2 values for Grouped Lasso (λ.min, fair comparison) tests were calculated for each of the three Y components:

```{r}
df_decomp_full <- df_full %>%
  mutate(
    natural_rate = pop_natural_change / pop_population * 100,
    social_rate  = pop_social_change  / pop_population * 100
  ) %>%
  filter(!is.na(natural_rate), !is.na(social_rate))

map_dfr(c(0, 1000, 5000, 10000), function(thr) {
  d <- df_decomp_full %>% filter(pop_population > thr)
  set.seed(1)
  ct <- sample(unique(d$city_code), 0.7 * length(unique(d$city_code)))
  gtr <- d %>% filter(city_code %in% ct)
  gte <- d %>% filter(!city_code %in% ct)
  foldid <- (as.integer(factor(gtr$city_code)) %% 10) + 1
  map_dfr(c("pop_growth_rate", "natural_rate", "social_rate"), function(yc) {
    cv  <- cv.glmnet(as.matrix(gtr[, fine_cols]),
                     gtr[[yc]],
                     alpha = 1,
                     foldid = foldid)
    pred <- as.numeric(predict(cv, as.matrix(gte[, fine_cols]), s = "lambda.min"))
    r <- suppressWarnings(cor(pred, gte[[yc]])^2)
    tibble(pop_min = thr,
           Y = yc,
           R2 = ifelse(is.na(r), 0, r))
  })
}) %>%
  pivot_wider(names_from = Y, values_from = R2) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

When pop \> 10,000, the R² values for the three components nearly double: total change from 0.127 to 0.251, natural change from 0.129 to 0.212, and social change from 0.064 to 0.156.

## Approach A: 4 Broad Categories of Linear Regression

```{r}
# Group the 18 variables into 4 broad categories based on their industrial significance (excluding “residential” as the reference category,to eliminate complete multicollinearity among the constituent data points)
df_prop <- df_prop %>% mutate(
  s_primary      = p_agri,
  s_secondary    = p_mining + p_plant + p_energy + p_it + p_logis,
  s_tertiary     = p_retail + p_finance + p_real_estate + p_hotel +
    p_education + p_health + p_other_service,
  s_public_other = p_official + p_other_type
)

df_train <- df_prop[idx, ]
df_test  <- df_prop[-idx, ]
```

```{r}
fitA <- lm(pop_growth_rate ~ s_primary + s_secondary + s_tertiary + s_public_other,
           data = df_train)
tidy(fitA) %>% mutate(across(where(is.numeric), ~ round(.x, 4)))
summary(fitA)$r.squared
```

### Explanation of Coefficients (Relative to the Residential Benchmark)

Since the sum of the four broad categories plus “residential” is always 1, once “residential” is excluded, each coefficient represents “the difference in the associated population growth rate when the share of construction shifts from residential to that industry” (correlation, not causation). In the main sample, all four coefficients are significantly negative:

- `s_primary` (Agriculture, Forestry, and Fisheries) −4.09, highest absolute value — High proportion of agricultural development, accompanied by lower population growth (characteristic of depopulation in rural areas)
- `s_public_other` (Public Affairs/Unclassified) −3.14 —— High proportion of development focused on remote administrative centers, accompanied by population decline
- `s_tertiary` (business services) −0.84, `s_secondary` (mining, energy, and transportation) −0.59 — also negative but weaker

The overall trend is very clear: the higher the share of residential construction, the higher the population growth rate; any industry that takes market share away from residential construction is associated with lower population growth. Compared to the full sample (where the coefficient is diluted by noise), the signal is stronger and purer in the main sample.

### Goodness of Fit and Diagnostics

```{r resid-A}
glance(fitA)[c("r.squared", "adj.r.squared")]

# Predicted values vs Residuals
augment(fitA) %>%
  ggplot(aes(.fitted, .resid)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted Values", y = "Residuals", title = "Residual plot (pop > 10,000)")
```

Building structures explain approximately 13.5% of the variance in population growth across actual municipalities.

### Test set

```{r}
pred_A <- predict(fitA, df_test)
r2_A   <- cor(pred_A, df_test$pop_growth_rate)^2
rmse_A <- sqrt(mean((pred_A - df_test$pop_growth_rate)^2))
c(R2 = r2_A, RMSE = rmse_A)
```

## Approach B: Fine-Grained Lasso

glmnet uses default normalization and L1 regularization for variable selection.

```{r cv-lasso}
X_train <- as.matrix(df_train[, fine_cols])
y_train <- df_train$pop_growth_rate
X_test  <- as.matrix(df_test[, fine_cols])
y_test  <- df_test$pop_growth_rate

set.seed(1)
cvfit <- cv.glmnet(X_train, y_train, alpha = 1)  # alpha=1 > lasso

cvfit$lambda.min
cvfit$lambda.1se

op <- par(mar = c(4, 4, 2, 1))  # smaller margins so it fits small panes
plot(cvfit)  # MSE vs log(lambda)
par(op)
```

### Industries Selected by the Sparse Model (lambda.1se)

```{r}
b_1se <- coef(cvfit, s = "lambda.1se")
selected <- data.frame(
  variable = rownames(b_1se),
  coef     = as.numeric(b_1se)
) %>%
  filter(variable != "(Intercept)", coef != 0) %>%
  arrange(coef) %>%
  mutate(coef = round(coef, 3))
selected

# Test set
pred_B_1se <- as.numeric(predict(cvfit, X_test, s = "lambda.1se"))
pred_B_min <- as.numeric(predict(cvfit, X_test, s = "lambda.min"))
r2_B_1se <- cor(pred_B_1se, y_test)^2
r2_B_min <- cor(pred_B_min, y_test)^2
c(R2_lambda1se = r2_B_1se, R2_lambda_min = r2_B_min)
```

## Analysis

Under the main sample, `lambda.1se` selected 13 industries, which naturally fell into two groups, fully aligning with the broad categories of Approach A:

- Positive correlation (population growth): `real_estate` (+4.29), `mixed_use` (+2.52), `logis` (+0.60) — real estate/mixed-use residential and commercial/logistics, corresponding to development corridors and population inflows
- Negative correlation (population outflow): `mining` (−3.88), `agri` (−2.99), `other_type` (−2.30), `official` (−1.77), etc. — dominated by mining, agriculture, and public services, corresponding to contraction in rural and remote areas

This indicates that the integration of domain knowledge (manual PCA) identified the correct direction, but the algorithm further distinguished, at the subcategory level, the few industries that truly drive the trend—particularly the strong positive signal from real estate and mixed-use residential and commercial development.

## Assessment of Credibility Tests: Grouped Cross-Validation

Data spanning five years for the same municipality can be used simultaneously for both training and testing. The model only needs to "remember" the fixed characteristics of a given municipality to "predict" its values for other years.

Truly meaningful generalization problem is "whether it can predict values for municipalities it has never seen before" which requires grouping by `city_code`.

### Group CV

```{r}
# Group by city_code
set.seed(1)
cities   <- unique(df_prop$city_code)
ct_train <- sample(cities, 0.7 * length(cities))
gtr <- df_prop %>% filter(city_code %in% ct_train)
gte <- df_prop %>% filter(!city_code %in% ct_train)
c(
  n_city_train = length(ct_train),
  n_city_test  = length(cities) - length(ct_train),
  n_row_train  = nrow(gtr),
  n_row_test = nrow(gte)
)

# Approach A
fitA_g  <- lm(pop_growth_rate ~ s_primary + s_secondary + s_tertiary + s_public_other,
              data = gtr)
r2_A_grp <- cor(predict(fitA_g, gte), gte$pop_growth_rate)^2

# Approach B

foldid_g <- (as.integer(factor(gtr$city_code)) %% 10) + 1
cvfit_g  <- cv.glmnet(as.matrix(gtr[, fine_cols]),
                      gtr$pop_growth_rate,
                      alpha = 1,
                      foldid = foldid_g)
pred_B_g <- as.numeric(predict(cvfit_g, as.matrix(gte[, fine_cols]), s = "lambda.1se"))
r2_B_grp <- cor(pred_B_g, gte$pop_growth_rate)^2

# Group selected by lasso
b_g <- coef(cvfit_g, s = "lambda.1se")
data.frame(variable = rownames(b_g), coef = as.numeric(b_g)) %>%
  filter(variable != "(Intercept)", coef != 0) %>%
  arrange(coef) %>% mutate(coef = round(coef, 3))

# Comparison
tibble(
  model = c("Approach A (Broad Categories)", "Approach B (lasso λ.1se)"),
  random_R2 = round(c(r2_A, r2_B_1se), 4),
  group_R2 = round(c(r2_A_grp, r2_B_grp), 4)
)
```

- Approach A：Random 0.104 \< Grouped 0.134 (Grouped is higher)
- Approach B：Random 0.173 \< Grouped 0.211 (Grouped is higher)

Contrary to expectations, the group-based R2 actually exceeded that of the random assignment, indicating that this portion of the signal represents genuine cross-municipality patterns rather than a memory of cross-year patterns within the same municipality.

Grouped lasso (λ=1se) selected 11 industries, showing a high degree of overlap with the 13 industries selected by random sampling `real_estate` (+3.36) and `mixed_use` (+2.98) remain the strongest positive performers, while `mining` (−3.45) and `agri` (−2.70) remain the strongest negative performers. Therefore, the conclusions are robust.

## Robustness Test for Time Lag

Population inflows drive new residential and logistics construction in the current year (reverse causality from Y to X), which exaggerates the correlation coefficient. To test whether the signal meets the temporal requirement that “X precedes Y,” replace Y with the population growth rate for the following year: use the building stock in year t to predict growth in year $t+1$.

```{r}
# Construct the population growth rate for year t+1, aligned by (city_code, year) back to year t
y_future <- df_prop %>%
  select(city_code, year, pop_growth_rate_future = pop_growth_rate) %>%
  mutate(year = year - 1)   # map to previous t+1

df_lag <- df_prop %>%
  left_join(y_future, by = c("city_code", "year")) %>%
  filter(!is.na(pop_growth_rate_future))

df_lag_train <- df_lag %>% filter(city_code %in% ct_train)
df_lag_test  <- df_lag %>% filter(!city_code %in% ct_train)
foldid_lag   <- (as.integer(factor(df_lag_train$city_code)) %% 10) + 1

cvfit_lag <- cv.glmnet(
  as.matrix(df_lag_train[, fine_cols]),
  df_lag_train$pop_growth_rate_future,
  alpha = 1,
  foldid = foldid_lag
)

b_lag <- coef(cvfit_lag, s = "lambda.1se")
selected_lag <- data.frame(
  variable = rownames(b_lag),
  coef     = as.numeric(b_lag)
) %>%
  filter(variable != "(Intercept)", coef != 0) %>%
  arrange(coef) %>%
  mutate(coef = round(coef, 3))
selected_lag

# R2 for the test set, compared with the group-based model
pred_lag <- as.numeric(predict(cvfit_lag,
  as.matrix(df_lag_test[, fine_cols]), s = "lambda.1se"))
r2_lag <- cor(pred_lag, df_lag_test$pop_growth_rate_future)^2
c(Same_Period_R2 = r2_B_grp, Lag_Period_R2 = r2_lag)
```

## Feature Engineering：Per Capita Construction

The previous model uses only the share of area (`p_*`), which captures industrial structure but ignores the scale of construction. Municipalities with the same industrial structure will clearly exhibit different population dynamics if their total construction volume is three times that of others. We supplement this by using per capita construction intensity:

$$ \log\!\Big(1 + \frac{\text{Total Area of New Building}}{\text{population}}\Big)$$

It is conceptually orthogonal to proportional characteristics—the sum of the proportions is always 1, regardless of the total amount.

```{r}
add_intensity <- . %>% mutate(
  total_area_per_cap = all_type_building_area / pop_population,
  log_area_per_cap   = log1p(total_area_per_cap)   
)
df_prop      <- add_intensity(df_prop)
gtr          <- add_intensity(gtr)
gte          <- add_intensity(gte)
df_lag_train <- add_intensity(df_lag_train)
df_lag_test  <- add_intensity(df_lag_test)
```

### Feature Evaluation

```{r intensity-scatter}
cor(df_prop$log_area_per_cap, df_prop$pop_population)

cor(df_prop$log_area_per_cap, df_prop$pop_growth_rate)

df_prop %>%
  select(log_area_per_cap, total_area_per_cap, pop_growth_rate) %>%
  pivot_longer(-pop_growth_rate, names_to="var", values_to="x") %>%
  mutate(var = if_else(var=="log_area_per_cap", "Logarithmic Intensity", "Initial Strength(square meters/person)")) %>%
  ggplot(aes(x, pop_growth_rate)) +
  geom_point(alpha=0.12, size=0.5) +
  geom_smooth(method="lm", se=FALSE, color="firebrick") +
  facet_wrap(~var, scales="free_x") +
  labs(x=NULL, y="Population Growth (%)",
       title="Per Capita Construction vs Population Growth (Initial vs Logarithmic)")
```

1.  Logarithmic intensity shows almost no correlation with population size (r ≈ 0.08)
2.  Logarithmic intensity is correlated with the population growth or decline rate: r = 0.41

### Feature Performance

```{r intensity-model}
# Approach A
fitA_int <- lm(pop_growth_rate ~ s_primary + s_secondary + s_tertiary +
                                 s_public_other + log_area_per_cap, data = gtr)
r2_A_int <- cor(predict(fitA_int, gte), gte$pop_growth_rate)^2

# Approach B
cvfit_int <- cv.glmnet(as.matrix(gtr[, c(fine_cols, "log_area_per_cap")]),
                       gtr$pop_growth_rate, alpha = 1, foldid = foldid_g)
pred_int <- as.numeric(predict(cvfit_int, as.matrix(gte[, c(fine_cols, "log_area_per_cap")]),
                               s = "lambda.1se"))
r2_B_int <- cor(pred_int, gte$pop_growth_rate)^2

tibble(
  Model = c("Approach A: Only proportion", "Approach A: Plus Capita",
           "Approach B: Only proportion", "Approach B: Plus Capita"),
  Group_R2 = round(c(r2_A_grp, r2_A_int, r2_B_grp, r2_B_int), 4)
)

b_int <- coef(cvfit_int, s = "lambda.1se")
data.frame(variable = rownames(b_int), coef = as.numeric(b_int)) %>%
  filter(variable == "log_area_per_cap") %>% mutate(coef = round(coef, 3))
```

The strength features doubled the R2 values for both sets of models: Approach A went from 0.134 to 0.44, and Approach B went from 0.211 to 0.426.

### Test for Time Lag (Approach B)

```{r}
# Approach B
foldid_il <- (as.integer(factor(df_lag_train$city_code)) %% 10) + 1

cvfit_lag_int <- cv.glmnet(
  as.matrix(df_lag_train[, c(fine_cols, "log_area_per_cap")]),
  df_lag_train$pop_growth_rate_future, alpha = 1, foldid = foldid_il
)
pred_lag_int <- as.numeric(predict(cvfit_lag_int,
  as.matrix(df_lag_test[, c(fine_cols, "log_area_per_cap")]), s = "lambda.1se"))
r2_lag_int <- cor(pred_lag_int, df_lag_test$pop_growth_rate_future)^2

tibble(
  Model = c("Same Period (X_t, Y_t)", "Lag (X_t, Y_t+1)"),
  `Only proportion` = round(c(r2_B_grp, r2_lag), 4),
  `Plus Capita` = round(c(r2_B_int, r2_lag_int), 4)
)
```

The lagged R2 (0.436) not only did not decrease but was actually slightly higher than the R2 (0.426) for the same period.
