---
title: "02-modeling"
author: "YANKAI MA"
date: "2026-07-22"
output: 
  pdf_document: 
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)   # lasso
```

## Data Preparation

```{r}


df <- read_csv(
  "data/cleaned_panel_2021_2025.csv",
  col_types = cols(city_code = col_character(), .default = col_guess()),
  show_col_types = FALSE
)
ncol_bldg <- ncol(df) - 6
colnames(df)[4:ncol_bldg] <- c(
  "all_type_building_count_total",
  "all_type_building_area",
  "living_count",
  "living_area",
  "dormitory_count",
  "dormitory_area",
  "mixed_use_count",
  "mixed_use_area",
  "agri_count",
  "agri_area",
  "mining_count",
  "mining_area",
  "plant_count",
  "plant_area",
  "energy_count",
  "energy_area",
  "it_count",
  "it_area",
  "logis_count",
  "logis_area",
  "retail_count",
  "retail_area",
  "finance_count",
  "finance_area",
  "real_estate_count",
  "real_estate_area",
  "hotel_count",
  "hotel_area",
  "education_count",
  "education_area",
  "health_count",
  "health_area",
  "other_service_count",
  "other_service_area",
  "official_count",
  "official_area",
  "other_type_count",
  "other_type_area"
)


# Percentage of total area accounted for by the 18 subcategories p
df_prop <- df %>%
  filter(all_type_building_area > 0) %>%            # Drop all zero rows
  mutate(
    across(
      ends_with("_area") & !all_type_building_area,
      ~ .x / all_type_building_area,
      .names = "p_{.col}"
    )
  ) %>%
  rename_with(~ str_remove(., "_area$"), starts_with("p_"))

# Proportions of the 17 subcategories (excluding p_living)
fine_cols <- setdiff(grep("^p_", names(df_prop), value = TRUE), "p_living")
# Full samples
df_full <- df_prop
# Threshold: pop > 10000
df_prop <- df_prop %>% filter(pop_population > 10000)

# 70/30 Split train and test set
set.seed(1)
idx <- sample(nrow(df_prop), 0.7 * nrow(df_prop))
df_train <- df_prop[idx, ]
df_test  <- df_prop[-idx, ]

```

## Explanation for `pop > 10000`

The EDA phase revealed that small villages and towns were outliers and were therefore excluded. The table below directly illustrates the benefits of this approach—after filtering by population threshold, the R2 values for Grouped Lasso (λ.min, fair comparison) tests were calculated for each of the three Y components:

```{r}
df_decomp_full <- df_full %>%
  mutate(
    natural_rate = pop_natural_change / pop_population * 100,
    social_rate  = pop_social_change  / pop_population * 100
  ) %>%
  filter(!is.na(natural_rate), !is.na(social_rate))

map_dfr(c(0, 1000, 5000, 10000), function(thr) {
  d <- df_decomp_full %>% filter(pop_population > thr)
  set.seed(1)
  ct <- sample(unique(d$city_code), 0.7 * length(unique(d$city_code)))
  gtr <- d %>% filter(city_code %in% ct)
  gte <- d %>% filter(!city_code %in% ct)
  foldid <- (as.integer(factor(gtr$city_code)) %% 10) + 1
  map_dfr(c("pop_growth_rate", "natural_rate", "social_rate"), function(yc) {
    cv  <- cv.glmnet(as.matrix(gtr[, fine_cols]),
                     gtr[[yc]],
                     alpha = 1,
                     foldid = foldid)
    pred <- as.numeric(predict(cv, as.matrix(gte[, fine_cols]), s = "lambda.min"))
    r <- suppressWarnings(cor(pred, gte[[yc]])^2)
    tibble(pop_min = thr,
           Y = yc,
           R2 = ifelse(is.na(r), 0, r))
  })
}) %>%
  pivot_wider(names_from = Y, values_from = R2) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

When pop \> 10,000, the R² values for the three components nearly double: total change from 0.127 to 0.251, natural change from 0.129 to 0.212, and social change from 0.064 to 0.156.
